Precise spiking motifs in neurobiological and neuromorphic data

### 2022-10-10 manuscript.tex
1.3

Indeed, the response of one biological neuron largely depends on the precise timing of the sequence of incoming spikes as they reach its soma. This \emph{event-based representation} present in the neuronal code is essential in understanding information processing and yet, most neuronal models do not take advantage of this minute temporal dimension. This hypothesis is directly inspired by neurobiological observations, particularly in the hippocampus, and it expands the capabilities of analog representations (which are based on the firing rate) by considering a representation based on repetitions of spiking motifs at precise times of occurrence. A mathematical formalization would be particularly well suited to neuromorphic computing, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Importantly, validating this hypothesis would also be crucial in our understanding of neuroscientific processes.

@@

In particular, there has been some remarkable findings when studying the dynamics of vision. Indeed, Simon Thorpe's group has shown during the last decades numerous examples demonstrating that humans can categorize briefly presented images in a fraction of a second. Its first experiment consisted in asking subjects to categorize images that do or do not contain animals~\citep{thorpe_speed_1996}. The results showed that humans were able to perform this task very well (with a success rate of more than 95\%) but above all that a differential activity for the two categories of images could be observed by electroencephalography, showing that this differentiation emerges at a very early latency in neuronal activity. These results have been extended to several species including primates but also extended to different experimental protocols and have shown for example that the response could be extremely fast of the order of 120 ms when the task was to perform a saccade~\citep{kirchner_ultra-rapid_2006}.

This fast processing also explains the surprising experiments of fast serial detection which consists in presenting a fast succession of different images and to decode via the EEG if the observer can detect for example the presence of an animal. The performances decrease progressively as the frequency of presentation of the images increases. However, it has been shown in the macaque that a significant performance could be maintained with an image presentation time of only 14 ms per image~\citep{keysers_speed_2001}.

This speed of the visual cortex, although surprising, is quite compatible with the latencies that are recorded at the neuro-physiological level. Indeed, when an image is presented on the retina, the visual information is rapidly propagated to the thalamus and then to the primary visual cortex takes about 55 ms in the macaque~\citep{nowak_timing_1997}, see Figure~\ref{fig:thorpe}. This functioning of visual processing as a forward pass is most prominent in fast processing but can be complemented with feedback loops from the higher areas to the sensory areas~\citep{lamme_distinct_2000}.

\begin{figure}
\centering
\note{get the rights or redo a schematics}
\includegraphics[width=.7\textwidth]{figures/visual-latency-estimate.jpg}
\caption{From~\citep{thorpe_seeking_2001} }\label{fig:thorpe}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{How timing encodes analogous profile}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

An important consequence of the speed of processing is that it implies that it is carried out using only very few spikes. Indeed, if we consider that a behavioral response in only 120 ms consists of about ten processing stages following the ``forward'' pathways of the visual system, then this imposes that the processing in a single area is performed with a reduced number of spikes. Neurophysiologists typically use firing sequences to characterize the activity of neurons using different statistics on their individual timing~\citep{perkel_neuronal_1967} but also the dependence across neurons~\citep{perkel_neuronal_1967-1}.

Numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and suggest the existence of repeated spatio-temporal patterns in the spike train. Note also that depending on how an in vitro cell is driven, this may influence the reliability of spike timing in neocortical neurons~\citep{mainen_reliability_1995}.

%----------------------------%
\begin{figure}
\centering
\note{redo a figure similar to the thorpe code}%\includegraphics{images/roc.jpg}
\caption{Latency coding. An input analog profile is encoded in latencies: the higher the contrast, the shorter the latency. In this example, one generates at most one spike per neuron.
}\label{fig:roc}
\end{figure}
%----------------------------%
At the level of dynamic processing of visual information, it has been shown that an encoding of the values of luminance imminence in the image instead of the retina~\citep{gollisch_rapid_2008}. Notably one can appreciate in figure 1 that the response of ganglion cells to visual gratings that are flashed onto the retina. The authors showed that the neuronal response could be encoded in the latency of the response and not only in the frequency of discharge as is often assumed. In figure 4 of the same article, these results are extended to natural images and show a qualitatively similar behavior. The conclusion of the authors is that the discharge latency of the neurons allows to encode spatial characteristics of the image. First-spike latency codes are a feasible mechanism for information transfer even when biologically plausible estimates of stimulus onset are taken into account~\citep{chase_first-spike_2007}. Interestingly, such fine latency effect can be at the origin of some visual illusions, for instance the illusion of false colors in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based based on center-surround interactions in the parvocellular pathway~\citep{kenyon_theory_2004}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Precise spike timings in neural assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{From synfire chains\ldots{}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate is highly irregular~\citep{softky_highly_1993}, which makes it inconsistent with temporal integration of random EPSPs. It has also been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible, where the first spike is aligned but the subsequent spike times tend to diffuse for independent repetitions of the stimulation~\citep{mainen_reliability_1995} (see Figur
~\ref{fig:mainen}). When driven by a highly dynamic \emph{frozen} noise, the output spikes are highly reproducible. This is consistent with the differential role of different stimulus frequencies on the reliability of spike timing reported in~\citep{nowak_influence_1997} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.''
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=.7\textwidth]{figures/MainenSejnowski1995.png} % TODO replace by the simulation
\caption{The timing of the spikes produced following the repetition of a step stimuli is less reproducible than that to a noisy stimulus. While this seems paradoxical at first sight, this is due to the use of the same \emph{frozen} noise at each repetition and highlight the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this \href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995.ipynb}{notebook} for a replication of this result using a simple LIF model.}\label{fig:mainen}
\end{figure}
%----------------------------%

>>>>

Indeed, the response of one biological neuron largely depends on the precise timing of the sequence of incoming spikes as they reach its soma. This \emph{event-based representation} present in the neuronal code is essential in understanding information processing and yet, most neuronal models do not take advantage of this minute temporal dimension. This hypothesis is directly inspired by neurobiological observations, particularly in the hippocampus, and it expands the capabilities of analog representations (which are based on the firing rate) by considering a representation based on repetitions of spiking motifs at precise times of occurrence. 
Additionally, numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and suggest the existence of repeated spatio-temporal patterns in the spike train. Note also that depending on how an in vitro cell is driven, this may influence the reliability of spike timing in neocortical neurons~\citep{mainen_reliability_1995}.  A mathematical formalization would be particularly well suited to neuromorphic computing, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Importantly, validating this hypothesis would also be crucial in our understanding of neuroscientific processes.

@@

In particular, there has been some remarkable findings when studying the dynamics of vision. Indeed, Simon Thorpe's group has shown during the last decades numerous examples demonstrating that humans can categorize briefly presented images in a fraction of a second. Its first experiment consisted in asking subjects to categorize images that do or do not contain animals~\citep{thorpe_speed_1996}. The results showed that humans were able to perform this task very well (with a success rate of more than 95\%) but above all that a differential activity for the two categories of images could be observed by electroencephalography, showing that this differentiation emerges at a very early latency in neuronal activity. These results have been extended to several species including primates but also extended to different experimental protocols and have shown for example that the response could be extremely fast of the order of 120 ms when the task was to perform a saccade~\citep{kirchner_ultra-rapid_2006}.
This fast processing also explains the surprising experiments of fast serial detection which consists in presenting a fast succession of different images and to decode via the EEG if the observer can detect for example the presence of an animal. The performances decrease progressively as the frequency of presentation of the images increases. However, it has been shown in the macaque that a significant performance could be maintained with an image presentation time of only 14 ms per image~\citep{keysers_speed_2001}.

This speed of the visual cortex, although surprising, is quite compatible with the latencies that are recorded at the neuro-physiological level. Indeed, when an image is presented on the retina, the visual information is rapidly propagated to the thalamus and then to the primary visual cortex takes about 55 ms in the macaque~\citep{nowak_timing_1997}, see Figure~\ref{fig:thorpe}. This functioning of visual processing as a forward pass is most prominent in fast processing but can be complemented with feedback loops from the higher areas to the sensory areas~\citep{lamme_distinct_2000}.
%
An important consequence of the speed of processing of vision is that it implies that it is carried out using only very few spikes. Indeed, if we consider that a behavioral response in only 120 ms consists of about ten processing stages following the ``forward'' pathways of the visual system, then this imposes that the processing in a single area is performed with a reduced number of spikes. Neurophysiologists typically use firing sequences to characterize the activity of neurons using different statistics on their individual timing~\citep{perkel_neuronal_1967} but also the dependence across neurons~\citep{perkel_neuronal_1967-1}.

%----------------------------%
\begin{figure}
\centering
\note{get the rights or redo a schematics}
\includegraphics[width=.7\textwidth]{figures/visual-latency-estimate.jpg}
\caption{From~\citep{thorpe_seeking_2001} }\label{fig:thorpe}
\end{figure}
%----------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{How timing encodes analogous profile}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%----------------------------%
\begin{figure}
\centering
\note{redo a figure similar to the thorpe code}%\includegraphics{images/roc.jpg}
\caption{Latency coding. An input analog profile is encoded in latencies: the higher the contrast, the shorter the latency. In this example, one generates at most one spike per neuron.
}\label{fig:roc}
\end{figure}
%----------------------------%
%
At the level of dynamic processing of visual information, it has been shown in the retina that an encoding of the values of luminance in the image in the timing of the spikes rather than the frequency~\citep{gollisch_rapid_2008}. Notably one can appreciate in these results that the response of ganglion cells to visual gratings that are flashed onto the retina. The authors showed that the neuronal response could be encoded in the latency of the response and not only in the frequency of discharge as is often assumed. These results were extended to natural images and show a qualitatively similar behavior. The conclusion of the authors is that the spiking latency of the neurons allows to encode spatial characteristics of the image. First-spike latency codes are a feasible mechanism for information transfer even when biologically plausible estimates of stimulus onset are taken into account~\citep{chase_first-spike_2007}. Interestingly, such fine latency effect can be at the origin of some visual illusions, for instance the illusion of false colors in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based  on center-surround interactions in the parvocellular pathway~\citep{kenyon_theory_2004}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Precise spike timings in neural assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{From synfire chains\ldots{}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate is highly irregular~\citep{softky_highly_1993}, which makes it inconsistent with temporal integration of random EPSPs. It has also been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible, where the first spike is aligned but the subsequent spike times tend to diffuse for independent repetitions of the stimulation~\citep{mainen_reliability_1995} (see Figur
~\ref{fig:mainen}). When driven by a highly dynamic \emph{frozen} noise, the output spikes are highly reproducible. This is consistent with the differential role of different stimulus frequencies on the reliability of spike timing reported in~\citep{nowak_influence_1997} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.''
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=.7\textwidth]{figures/MainenSejnowski1995.png} % TODO replace by the simulation
\caption{The timing of the spikes produced following the repetition of a step stimuli is less reproducible than that to a noisy stimulus. While this seems paradoxical at first sight, this is due to the use of the same \emph{frozen} noise at each repetition and highlight the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this \href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995.ipynb}{notebook} for a replication of this result using a simple LIF model.}\label{fig:mainen}
\end{figure}
%----------------------------%
